---
title: "dlmextra - introduction"
author: "Søren Højsgaard"
date: "`r date()`"
output: 
  rmarkdown::html_vignette: 
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{dlmextra - introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
options("digits"=3, "warn"=-1)
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The `dlmextra` package (located at
[https://github.com/hojsgaard/dlmextra]) enhances the `dlm` package
(see [https://cran.r-project.org/web/packages/dlm/index.html]) by
providing extra functionality. Functions from `dlm` are indicated with
`::`, e.g. `dlm::dlmModReg`. Functions from `dlmextra` are not
indicated in any special way. Thereby it should be possible to see
what comes from where.
Most issues adressed in `dlmextra` relate to models with time varying
components. Please do report back on problems with `dlmextra`.

```{r, message=FALSE}
library(dlmextra)
library(ggplot2)
```

# A small example: A dynamic regression model

```{r}
tt   <- seq(1, 10, by=1)
beta <- c(1, -.2)
dat  <- data.frame(tt=tt)
M    <- model.matrix(~ tt, data=dat)
dat$mu <- M %*% beta
set.seed(112021)
dat$y0 <- as.numeric(dat$mu) + rnorm(length(dat$mu), sd=.5)
```

We have
```{r}
head(dat)
head(M)
```

```{r}
pt0 <- ggplot(data=dat) + geom_point(aes(tt, y0)) +
    geom_line(aes(x=tt, y=mu)) 
pt0
```

## Dynamic regression model in `dlm`

A dynamic regression model has the form
\begin{align}
 y_t &= F_t \theta_t + v_t;          & v_t \sim N(0, V_t)\\
 \theta_t &= G_t \theta_{t-1} + w_t; & w_t \sim N(0,W_t)
\end{align}
where the matrix $F_t=[1, t]$ is
time dependent whereas $G_t$, $V_t$ and $W_t$ are not
(notice that $G_t=I$). 

```{r}
drm1 <- dlm::dlmModReg(M, addInt=FALSE, m0=beta, C0=diag(1,2), dW=rep(.1, 2))
drm1
```

Several issues arise with models containing time dependent
components. (These issues do not appear if a model only contains
constant components).

## Accessing a model object

The $F_t$ matrix is constructed on the fly on the basis of
the model matrix and the specifications in `JFF`:

```{r}
dlm::FF(drm1)
dlm::JFF(drm1)
FFeval(drm1, 4) ## or
getdlm(drm1, "F", 4)
```

# Filtering

Imagine that we have only seen the first observations:
```{r}
nseen  <- 4
dat$y <- dat$y0
dat$y[-(1:nseen)] <- NA
dat$y
```

Filtering works because the time varying components can be constructed
on the fly because the number of rows in the design matrix is at least
the same as the length of $y$ (including the missing values in
$y$). We have seen this done above for $F_t$.

Notice that there are two underlying engines for the
filtering. Fortunately we get same results, but that is not always the
case as we shall see below.

```{r}
fm1C <- dlm::dlmFilter(dat$y, drm1)          ## Filtering based on C code; default
fm1R <- dlm::dlmFilter(dat$y, drm1, debug=T) ## Filtering based on R code
fm1C$f
fm1R$f
```


## Filtering vectors that are "too long" or "too short"

Consider filtering data vector $y$ vector with fewer elements than the
number of rows in the model matrix:

```{r}
yshort <- dat$y[1:5]
ylong <- c(dat$y, NA, NA)
```


```{r}
fm2C <- dlm::dlmFilter(yshort, drm1)          ## Using C code
fm2R <- dlm::dlmFilter(yshort, drm1, debug=T) ## Using R code
fm2C$f
fm2R$f
```
When filtering is based on R code, we get results consistent with what
happened when filtering the full $y$ vector; with C code we get
something else. This behaviour does not seem to be documented anywhere.

Next consider filtering a vector with more elements than the number of
rows in the model matrix.  (There is no information in the supplied
model matrix for generating $F_t$ for $t>6$). The R based engine
correctly captures the problem; the C based engine does not. This
behaviour does not seem to be documented anywhere.

```{r}
fm3C <- try(dlm::dlmFilter(ylong, drm1))          ## Using C code
fm3R <- try(dlm::dlmFilter(ylong, drm1, debug=T)) ## Using R code
names(fm3C)
fm3R
```

## Recommendation: use `dlmxFilter`

Based on these findings, I recommend to stay on the safe side and
always set `debug=T`. (Fast C code is nice; correct results are
nicer.) In `dlmextra`, the `dlmxFilter` function only uses the R code,
so sticking to that function should make you safe.


## Accessing a filtered object

```{r}
names(fm1R)
```

As shown above, a filtered object contains data `y`, the model object
`mod` itself, the slots `m`, `a`, `f` (refer to the book for these
quantities) but not `R` and `C`. However, these quantities can be
constructed / extracted as:

```{r}
sel <- 1:2
getdlm(fm1R, "m", select=sel)
getdlm(fm1R, "sdC", select=sel) # Sqrt of diagonals of C
getdlm(fm1R, "a", select=sel)
getdlm(fm1R, "sdR", select=sel) # Sqrt of diagonals of R
getdlm(fm1R, "f", sel=sel)
```

List of matrices with $G$ and $R$:
```{r}
getdlm(fm1R, "C", select=sel)
getdlm(fm1R, "R", select=sel)
```


# Estimation

The issue about using R or C code also arises in connection with
estimation. Say we want to estimate the unknown variances:

```{r}
V(drm1)
W(drm1)
```

The way ahead is along these lines. Notice that the number of rows in
the model matrix is the same as the number of elements in $y$. Notice
that there are two underlying engines for estimation. Fortunately
we get same results, but that is not always the case as we shall see
below.


```{r}
build_obj <- function(p, obj){
    V(obj) <- exp(p[1])
    diag(W(obj)) <- exp(p[2:3])
    obj
}

p <- c(0.1, 0.2, 0.3)

fit1C <- dlmMLE(dat$y, parm=p, build=build_obj, obj=drm1)          ## Using C code
fit1R <- dlmMLE(dat$y, parm=p, build=build_obj, obj=drm1, debug=T) ## Using R code
fit1C$par
fit1R$par
```

## Estimation with vectors that are "too long" or "too short"

Consider estimation with at data vector $y$ vector with fewer elements than the
number of rows in the model matrix:

```{r}
fit2C <- dlmMLE(yshort, parm=p, build=build_obj, obj=drm1) ## Wrong!
fit2R <- dlmMLE(yshort, parm=p, build=build_obj, obj=drm1, debug=T)
fit2C$par
fit2R$par
```

When estimation is based on R code, we get results consistent with what
happened when estimating with the full y vector; with C code we get
something else. This behaviour does not seem to be documented
anywhere.

Next consider estimation with a vector with more elements than the number of
rows in the model matrix.  (There is no information in the supplied
model matrix for generating $F_t$ for $t>6$). The R based engine
correctly captures the problem; the C based engine does not. This
behaviour does not seem to be documented anywhere.


```{r}
fit3C <- try(dlmMLE(ylong, parm=p, build=build_obj, obj=drm1))
fit3R <- try(dlmMLE(ylong, parm=p, build=build_obj, obj=drm1, debug=T))

fit3C$par
fit3R
```

## Recommendation: use `dlmxMLE`

Based on these findings, I recommend to stay on the safe side and
always set `debug=T`. (Fast C code is nice; correct results are
nicer.) In `dlmextra`, the `dlmxMLE` function only uses the R code, so
sticking to that function should make you safe. The `dlmxMLE` function
also check that dimension of $y$ and the model matrix match.

One feature of `dlmxMLE` is that the function returns the model object
with the estimated parameters.

```{r}
xx <- dlmxMLE(yshort, parm=p, build=build_obj, obj=drm1)
xx$par
exp(xx$par)
zz <- xx$mod
V(zz)
W(zz)
```













# Forecasting

The `dlm::dlmForecast` function does not work with non-constant
models, i.e. models where $F$, $G$, $V$ or $W$ are vary with time. The
problem is that if one of the components is time varying then this
compenent needs to be constructed on the fly as explained above.

```{r}
try(dlm::dlmForecast(fm1R, nAhead=length(tt) - nseen))
```

## Recommendation: use `dlmxForecast`

However, forecasting is nothing but filtering with missing
data. Hence, as long as the model matrix in the model object has
"enough rows" for constructing time dependent quantities we are fine
and can make the necessary computations as illustrated above:

```{r, eval=T}
v  <- dlmxForecast(fm1R, nAhead=length(tt)-nseen, offset=nseen)
ci <- confint(v)
ci
```

# A couple of nice plots

Now we can plot with prediction intervals etc.
```{r}
ci$tt <- tt[ci$xrow]
pt1 <- pt0 +
    geom_vline(xintercept=tt[nseen], col="red")   + 
    geom_line(aes(x=tt, y=fit), data=ci, col="purple") +
    geom_errorbar(aes(x=tt, ymin=lwr, ymax=upr), data=ci)
pt1
```
