---
title: "dlmextra - introduction"
author: "Søren Højsgaard"
date: "`r date()`"
output: 
  rmarkdown::html_vignette: 
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{dlmextra - introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
options("digits"=3, "warn"=-1)
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The `dlmextra` package (located at
[https://github.com/hojsgaard/dlmextra]) enhances the `dlm` package
(see [https://cran.r-project.org/web/packages/dlm/index.html]) by
providing extra functionality.

```{r, message=FALSE}
library(dlmextra)
library(ggplot2)
```

# A small example: A dynamic regression model

```{r}
tt   <- seq(1, 6, by=1)
beta <- c(1, -.2)
dat  <- data.frame(tt=tt)
M    <- model.matrix(~ tt, data=dat)
dat$mu <- M %*% beta
set.seed(112021)
dat$y0 <- as.numeric(dat$mu) + rnorm(length(dat$mu), sd=.5)
```

```{r}
pt0 <- ggplot(data=dat) + geom_point(aes(tt, y0)) +
    geom_line(aes(x=tt, y=mu)) 
pt0
```

## Dynamic regression model in `dlm`

```{r}
mm1 <- dlmModReg(M, addInt=FALSE, m0=beta, C0=diag(1,2), dW=rep(.1, 2))
mm1
```

The model is

\begin{align}
 y_t &= F_t \theta_t + v_t;          & v_t \sim N(0, V_t)\\
 \theta_t &= G_t \theta_{t-1} + w_t; & w_t \sim N(0,W_t)
\end{align}

As the model is specified in `dlmModReg` 
the  matrix $F_t=[1, t]$ is really time
dependent whereas $G_t$, $V_t$ and $W_t$ are not ($G_t=I$). 
The $F_t$ matrix is
constructed on the fly on the basis of the model matrix:

```{r}
getFF(mm1, 1)
getFF(mm1, 2)
```


Imagine that we have only seen the first observations:
```{r}
nseen  <- 3
dat$y <- dat$y0
dat$y[-(1:nseen)] <- NA
dat$y
```

We have
```{r}
dat
M
```


## Filtering

Filtering works because the time varying components can be constructed
on the fly because the number of rows in the design matrix is at least
the same as the length of $y$ (including the missing values in $y$).
Notice that there are two underlying engines for the
filtering. Fortunately we get same results, but that is not always the
case as we shall see below.

```{r}
fm1C <- dlmFilter(dat$y, mm1)          ## Filtering based on C code; default
fm1R <- dlmFilter(dat$y, mm1, debug=T) ## Filtering based on R code
fm1C$f
fm1R$f
## FIXME: summary(fm1) would be nice
```


### Important issue: Filtering a shorter vector

Consider filtering a shorter $y$ vector:

```{r}
yshort <- dat$y[1:4]
fm2C <- dlmFilter(yshort, mm1)          ## Using C code
fm2R <- dlmFilter(yshort, mm1, debug=T) ## Using R code
fm2C$f
fm2R$f
```

When filtering is based on R code, we get results consistent with what
happened when filtering the full $y$ vector; with C code we get
something else. This behaviour does not seem to be documented anywhere.

### Important issue: Filtering a longer vector

A related issue is the following: Filtering a $y$ vector longer than
the number of rows in the design matrix should give an error in a
non-constant model. (There is no information in the supplied model
matrix for generating $F_t$ for $t>6$). The R based engine correctly
captures the problem; the C based engine does not. This behaviour does
not seem to be documented anywhere.

```{r}
ylong <- c(dat$y, NA, NA)
fm3C <- try(dlmFilter(ylong, mm1))          ## Using C code
fm3R <- try(dlmFilter(ylong, mm1, debug=T)) ## Using R code
fm3C %>% lapply(head, 3)
fm3R
```


## Important issue: Recommendation

Based on these findings, I recommend to stay on the safe side and
always set `debug=T`. (Fast C code is nice; correct results
are nicer.)

```{r}
## FIXME: Should handle this in debug=T-issue in dlmextra
```

## Accessing a filtered object

```{r}
sel <- 1:6
getdlmf(fm1R, "m", select=sel)
getdlmf(fm1R, "C", select=sel)
getdlmf(fm1R, "a", select=sel)
getdlmf(fm1R, "R", select=sel)
## FIXME: diagCC and diagRR would be nice
```

## Filtering and forecasting

The `dlm::dlmForecast` function does not work with non-constant
models, i.e. models where $F$, $G$, $V$ or $W$ are vary with time. The
problem is that if one of the components is time varying then this
compenent needs to be constructed on the fly as explained above.

```{r}
try(dlm::dlmForecast(fm1R, nAhead=length(tt) - nseen))
```

However, forecasting is nothing but filtering with missing
data. Hence, as long as the model matrix in the model object has
"enough rows" for constructing time dependent quantities we are fine
and can make the necessary computations as illustrated above:

```{r, eval=T}
v  <- dlmXForecast(fm1R, nAhead=length(tt)-nseen, offset=nseen)
ci <- confint(v)
ci
```

## A couple of nice plots

Now we can plot with prediction intervals etc.
```{r}
ci$tt <- tt[ci$xrow]
pt1 <- pt0 +
    geom_vline(xintercept=tt[nseen], col="red")   + 
    geom_line(aes(x=tt, y=fit), data=ci, col="purple") +
    geom_errorbar(aes(x=tt, ymin=lwr, ymax=upr), data=ci)
pt1
```










