---
title: "`dlmextra` - Additional functionality to the `dlm` package"
author: "Søren Højsgaard"
date: "`r date()`"
output: 
  rmarkdown::html_vignette: 
    toc: true
vignette: >
  %\VignetteIndexEntry{extra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, include=FALSE}
options("digits"=3)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The `dlmextra` package enhances `dlm` by providing extra functionality
aimed at simplifying the usage of the `dlm` package.

```{r, message=FALSE}
library(dlmextra)
library(ggplot2)
```

## A small example: A dynamic regression model

```{r}
tt <- seq(1, 6, by=1)
beta <- c(1, -1)
dat  <- data.frame(tt=tt)
M    <- model.matrix(~ tt, data=dat)
dat$mu   <- M %*% beta

set.seed(121314)
dat$y0 <- as.numeric(dat$mu) + rnorm(length(dat$mu))
pt0 <- ggplot(data=dat) + geom_point(aes(tt, y0)) +
    geom_line(aes(x=tt, y=mu)) 
pt0
```

### Dynamic regression model

```{r}
mm1 <- dlmModReg(M, addInt=FALSE, m0=beta, C0=diag(1,2), dW=rep(.1, 2))
mm1
```


Now imagine that the observed data are:
```{r}
nseen  <- 3
dat$y <- dat$y0
dat$y[-(1:nseen)] <- NA
dat$y
```

### Filtering

Filtering works because the time varying components can be constructed
on the fly because the number of rows in the design matrix is at least
the same as the length of $y$ (including the missing values in $y$). 

Notice that
there are two underlying engines for the filtering: 

```{r}
fm1 <- dlmFilter(dat$y, mm1)          ## Filtering based on C code
fm1$f

fm1 <- dlmFilter(dat$y, mm1, debug=T) ## Filtering based on R code
fm1$f
## FIXME: summary(fm1) would be nice
```

Fortunately we get same results, but that is not always the
case. 


### Filtering a shorter vector

Consider filtering a shorter $y$ vector:

```{r}
yshort <- dat$y[1:4]
fm2 <- dlmFilter(yshort, mm1)          ## Using C code
fm2$f

fm2 <- dlmFilter(yshort, mm1, debug=T) ## Using R code
fm2$f
```

When filtering is based on R code, we get results consistent with what
happened when filtering the full $y$ vector; with C code we get
something else. This behaviour seems not to be documented anywhere.

### Filtering a longer vector

A related issue is the following: Filtering a $y$ vector longer than
the number of rows in the design matrix should give an error in a
non-constant model. The R based engine correctly captures the problem;
the C based engine does not.

```{r}
ylong <- c(dat$y, NA, NA)
fm2 <- try(dlmFilter(ylong, mm1))          ## Using C code
fm2$f

fm2 <- try(dlmFilter(ylong, mm1, debug=T)) ## Using R code
fm2
```



### Recommendation

Based on these findings, I recommend to stay on the safe side and
always set `debug=T`. (Fast C code is of course nice; correct results
are nicer.)

```{r}
## FIXME: Should handle this in debug=T-issue in dlmextra
```




```{r}
sel <- 1:4
mm(fm1, select=sel)
CC(fm1, select=sel)
aa(fm1, select=sel)
RR(fm1, select=sel)
## FIXME: diagCC and diagRR would be nice
```



## Filtering and forecasting

The `dlm::dlmForecast` function does not work with non-constant
models, i.e. models where $F$, $G$, $V$ or $W$ are vary with time. The
problem is that if one of the components is time varying then this
compenent needs to be constructed on the fly, and this applies also for the time points
for which forecasting is made.

```{r}
try(dlm::dlmForecast(fm1, nAhead=length(tt) - nseen))
```

However, forecasting is nothing but filtering with missing
data. Hence, as long as the model matrix in the model object has
enough rows we are fine and can make the necessary computations as
illustrated above:

```{r, eval=T}
v  <- dlmXForecast(fm1, nAhead=length(tt) - nseen, offset=nseen)
ci <- confint(v)
ci
```

Now we can plot with prediction intervals etc.
```{r}
ci$tt <- tt[ci$xrow]
pt1 <- pt0 +
    geom_vline(xintercept=tt[nseen], col="red")   + 
    geom_line(aes(x=tt, y=fit), data=ci, col="purple") +
    geom_errorbar(aes(x=tt, ymin=lwr, ymax=upr), data=ci)
pt1
```










